---
title: "Playing with Random Numbers"
author: "Based on script by Alexander Skupin"
date: "2025-10-21"
format: html
---

## Introduction

This Quarto document explores pseudo-random number generation (PRNG) and related distributions in R, relevant for simulations in bioinformatics such as modeling biological variability or Monte Carlo methods. We generate sequences, analyze properties like mean, variance, and entropy, and compare distributions using built-in R functions.

## Pseudo-Random Number Generation (PRNG)

PRNG uses a deterministic algorithm to produce sequences that mimic randomness. Here, a linear congruential generator is implemented with parameters: starting point (xs), multiplier (mul), and drift (dri).

```{r}
nn <- 4000  # number of sequence

# starting point
xs <- 1/123  # initial seed value

# multiplier 
mul <- 2.132  # scaling factor

# drift 
dri <- 0.212134  # additive constant

# vector for random numbers and initiation 
rannum <- rep(0, nn)
rannum[1] <- xs

# loop to generate RN
for (i in 2:nn) {
  rannum[i] <- (mul * rannum[i-1] + dri) %% 1
}
```

## Visualizing the Generated Numbers

Plot the sequence to inspect randomness visually. A uniform scatter suggests good randomness.

```{r}
plot(rannum)
```

Plot consecutive pairs to check for dependencies or patterns (e.g., correlations).

```{r}
plot(rannum[1:(nn-1)], rannum[2:nn])  # plot dependencies
```

## Histogram Analysis

Histograms reveal the distribution. For uniform PRNG, it should be flat across [0,1).

```{r}
hist(rannum, breaks = 60)  # histogram
```

Detailed histogram for further properties.

```{r}
disr <- hist(rannum, breaks = 250)  # properties of histogram
```

## Statistical Properties

Check mean (expected ~0.5 for uniform [0,1)) and variance (expected ~1/12).

```{r}
mean(rannum)  # average (to be 0.5 - see lecture)
sd(rannum)^2  # variance (to be 1/12 - see lecture)
```

## Entropy Calculation

Entropy measures information or uncertainty. For discrete bins, compute Shannon entropy.

```{r}
# calculate entropy and average
ent <- 0
danv <- 0
for (i in 1:length(disr$counts)) {
  ptemp <- disr$counts[i] / nn
  ent <- ent - ptemp * log2(ptemp)
  danv <- danv + ptemp * disr$mids[i]  # the mids are the x
}

print(ent)  # depends on binning
print(danv)
```

Note: For uniform distribution, analytical entropy is log(b-a) = 0, but binning affects computation.

## Differential Entropy

For continuous distributions, adjust for bin width in entropy calculation.

```{r}
# calculate differential entropy
nbis <- 500
ddisr <- hist(rannum, breaks = nbis)  # properties of histogram
dent <- 0
for (i in 1:length(ddisr$counts)) {
  ptemp <- ddisr$counts[i] / nn / nbis  # note: adjusted for bin width (1/nbis assuming range 1)
  if (ptemp > 0) { dent <- dent - ptemp * log2(ptemp) }
}

print(dent)  # depends on binning
```

## Exponential Distribution via Inverse Sampling

Transform uniform to exponential using inverse CDF: -λ log(u), where u ~ Uniform[0,1).

```{r}
lambda <- 1  # rate parameter
expran <- -lambda * log(rannum)  # inverse function of exponential
expden <- hist(expran, breaks = 500, freq = FALSE)
```

Check normalization and statistics.

```{r}
sum(expden$counts) / nn  # check if normalized
sum(expden$counts * expden$mids) / nn  # average by distribution
mean(expran)  # average of numbers
sd(expran)  # standard deviation
```

## Uniform Random Numbers with R's runif

R's built-in uniform generator for reproducibility and efficiency.

```{r}
runif(10)  # generate 10 uniform numbers
```

Plot sequence.

```{r}
plot(runif(1000))
```

Mean of large sample.

```{r}
mean(runif(100000))
```

Dependencies and histogram.

```{r}
rn <- runif(1000)
plot(rn[1:999], rn[2:1000])
rnhist <- hist(rn, breaks = 30)
```

Normalized histogram and density.

```{r}
sum(rnhist$counts / 1000)  # check if normalized
barplot(rnhist$counts / 1000)  # plot normalized histogram
plot(density(rn))
```

Effect of binning.

```{r}
rnhist <- hist(rn, breaks = 100)
rnhist <- hist(rn, breaks = 30)
```

## Scaled Uniform and Sum of Uniforms

Scale uniform and sum two for triangular distribution (like sum of dice).

```{r}
rn2 <- 2 * runif(1000)
plot(rn[1:100], ylim = c(0, 2))
points(rn2, col = "red")
```

Densities.

```{r}
plot(density(rn), xlim = c(0, 3))
lines(density(rn2), col = "red")
```

Sum.

```{r}
rnsum <- rn + rn2
hist(rnsum)
plot(density(rnsum), ylim = c(0, 2))
```

Compare densities.

```{r}
plot(density(rn), ylim = c(0, 2), xlim = c(0, 4))
lines(density(rn2), col = "red")
lines(density(rnsum), col = "blue")
```

## Exponential with runif

Direct generation.

```{r}
hist(-log(runif(100000)))
plot(density(-log(runif(1000000))))
```

As vector.

```{r}
rne <- -log(runif(1000))
rnehist <- hist(rne)
sum(rnehist$counts / 1000)  # check if normalized
barplot(rnehist$counts / 1000)  # plot normalized histogram
```

## Gaussian Random Numbers with rnorm

Normal distribution, common in bioinformatics for noise or measurements.

```{r}
rng <- rnorm(10000, mean = 10, sd = 2)  # with mean 10 and sd 2
plot(rng[1:1000])
rnghist <- hist(rng)
sum(rnghist$counts / 10000)  # check if normalized
barplot(rnghist$counts / 10000)  # plot normalized histogram
```

## Significance Tests

Compare distributions using non-parametric (Wilcoxon) and parametric (t-test) tests.

```{r}
wilcox.test(rng, rn)
t.test(rng, rne)  # assumes normality
```

Histograms for visual comparison.

```{r}
hist(rn)
hist(rng)
hist(rne)
```

## Reproducibility with set.seed

Set seed for identical results across runs.

```{r}
set.seed(2)
rng2 <- rnorm(1000, mean = 9.5, sd = 2)
plot(rng2[1:100])
points(rng, col = "red")
```

Densities.

```{r}
plot(density(rng))
lines(density(rng2), col = "red")
```

Tests with full and subsamples.

```{r}
wilcox.test(rng, rng2)
wilcox.test(rng[1:120], rng2[1:120])  # significance depends on sample size
t.test(rng, rng2)
t.test(rng[1:120], rng2[1:120])
```