---
title: "Single-Cell RNA-Seq Analysis"
author: "Dimitrios Kyriakis and Alexander Skupin"
affiliation: University of Luxembourg
date: "2025-10-21"
format: 
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Table of Contents"
    code-fold: true
    code-tools: true
    theme: cosmo
    css: styles.css
    number-sections: true
engine: knitr
filters:
  - webr
webr:
  show-editor: true
  show-startup-message: false
  autorun: true
number-sections: true
message: false
warning: false
---

## Introduction

This Quarto document demonstrates single-cell RNA-seq analysis using R packages `SingleCellExperiment`, `Seurat`, and `scater`. These tools are essential in bioinformatics for quality control (QC), normalization, dimensionality reduction, and clustering of single-cell transcriptomic data. We use the PBMC3k dataset to explore these methods.

## Loading Libraries and Data

Load required libraries and the PBMC3k dataset from 10X Genomics. Ensure the data is unzipped and placed in the correct directory.

```{r}
# Reset environment
rm(list=ls())

library(SingleCellExperiment)
library(dplyr)
library(Seurat)
library(patchwork)

# Download and unzip PBMC dataset from URL
url <- "https://s3-us-west-2.amazonaws.com/10x.files/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz"
temp_file <- tempfile(fileext = ".tar.gz")
download.file(url, temp_file, mode = "wb")
temp_dir <- tempdir()
untar(temp_file, exdir = temp_dir)
pbmc.data <- Read10X(data.dir = file.path(temp_dir, "filtered_gene_bc_matrices/hg19"))
```


## Initial Data Exploration

Examine the dataset dimensions and gene expression for specific genes in the first 30 cells.

```{r}
dim(pbmc.data)  # dimensions of data frame
head(pbmc.data@Dimnames[[1]])  # gene names
head(pbmc.data@Dimnames[[2]])  # barcode names
pbmc.data[c("CD3D", "TCL1A", "MS4A1"), 1:30]
```

Visualize gene expression distributions with histograms.

```{r}
hist(pbmc.data["CD3D",], breaks = 50)
hist(pbmc.data[122, 1:2230])
```

## Creating Seurat and SingleCellExperiment Objects

Initialize a Seurat object for analysis, filtering genes expressed in at least 3 cells and cells with at least 200 features.

```{r}
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)
pbmc.sce <- as.SingleCellExperiment(pbmc)
pbmc
```

## Quality Control (QC)

Calculate mitochondrial gene percentage for QC, as high mitochondrial content may indicate stressed or damaged cells.

```{r}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
head(pbmc@meta.data, 5)  # show QC metrics
hist(pbmc@meta.data$percent.mt, breaks = 50)
```

Visualize QC metrics with violin plots and scatter plots.

```{r}
vlnpl1 <- VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot3 <- plot1 + plot2
```

Filter cells based on feature counts (200–2500) and mitochondrial percentage (<5%).

```{r}
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
vlnpl2 <- VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
vlnpl1  # before QC
vlnpl2  # after QC
plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot4 <- plot1 + plot2
plot3  # compare with before QC
```

## Data Normalization

Normalize data using LogNormalize to account for sequencing depth.

```{r}
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 10000)
```

## Feature Selection

Identify the 2000 most variable genes for downstream analysis.

```{r}
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)
top10 <- head(VariableFeatures(pbmc), 10)
plot5 <- VariableFeaturePlot(pbmc)
```

Scale data for PCA, focusing on variable genes.

```{r}
pbmc <- ScaleData(pbmc)
```

Access metadata, counts, and specific genes.

```{r}
head(Features(pbmc))  # gene names
head(Cells(pbmc))  # cell barcodes
head(pbmc[[]])  # metadata
head(GetAssayData(pbmc, assay = "RNA", slot = "counts"))  # count matrix
head(FetchData(pbmc, vars = c("nFeature_RNA", "ISG15", "TNFRSF4"), layer = "counts"))
```

Visualize gene expression.

```{r}
plot(as.data.frame(FetchData(pbmc, vars = c("FTL"), layer = "counts"))[,])
plot(as.data.frame(FetchData(pbmc, vars = c("FTL"), layer = "counts"))[,],
     as.data.frame(FetchData(pbmc, vars = c("GAPDH"), layer = "counts"))[,])
summary(as.data.frame(FetchData(pbmc, vars = c("FTL"), layer = "counts")))
```

## Dimensionality Reduction

Perform PCA on scaled data to reduce dimensionality.

```{r}
pbmc <- RunPCA(pbmc, npcs = 30, verbose = FALSE)
ElbowPlot(pbmc, ndims = 30)
print(pbmc[["pca"]], dims = 1:5, nfeatures = 5)
DimPlot(pbmc, reduction = "pca")
VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
DimHeatmap(pbmc, dims = 1:15, cells = 500, balanced = TRUE)
```

Determine significant PCs using JackStraw.

```{r}
pbmc <- JackStraw(pbmc, num.replicate = 50)
pbmc <- ScoreJackStraw(pbmc, dims = 1:20)
JackStrawPlot(pbmc, dims = 1:10)
```

## Cell Cycle Scoring and Regression

Assign cell cycle phases and regress out cell cycle effects.

```{r}
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
pbmc <- CellCycleScoring(pbmc, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)
DimPlot(pbmc, group.by = "Phase")
pbmc$CC.Difference <- pbmc$S.Score - pbmc$G2M.Score
pbmc <- ScaleData(pbmc, vars.to.regress = "CC.Difference", features = rownames(pbmc))
pbmc <- RunPCA(pbmc, features = VariableFeatures(pbmc), reduction.name = "pca_reg_ccs")
DimPlot(pbmc, reduction = "pca_reg_ccs") | DimPlot(pbmc, reduction = "pca")
```

## Clustering

Cluster cells using the Louvain algorithm on PCA results.

```{r}
pbmc <- FindNeighbors(pbmc, reduction = "pca", dims = 1:20, verbose = FALSE) %>%
  FindClusters(resolution = 0.5, verbose = FALSE)
DimPlot(pbmc, group.by = "seurat_clusters")
```

## Non-Linear Dimensionality Reduction

Run UMAP to visualize clusters, noting the stochastic nature of UMAP.

```{r}
pbmc <- RunUMAP(pbmc, reduction = "pca", dims = 1:20, verbose = FALSE)
p1 <- DimPlot(pbmc, label = TRUE, repel = TRUE, reduction = "pca")
p2 <- DimPlot(pbmc, label = TRUE, repel = TRUE, reduction = "umap")
pbmc <- RunUMAP(pbmc, reduction = "pca", dims = 1:20, verbose = FALSE, seed.use = 1234)
p3 <- DimPlot(pbmc, label = TRUE, repel = TRUE, reduction = "umap")
p1 | p2 | p3
```

## Differential Expression Analysis

Identify markers for cluster 2 and all clusters.

```{r}
cluster2.markers <- FindMarkers(pbmc, ident.1 = 2, min.pct = 0.25, min.diff.pct = 0.4)
cluster2.markers$Symbol <- rownames(cluster2.markers)
head(cluster2.markers, n = 5)
VlnPlot(pbmc, features = cluster2.markers$Symbol[1:2])
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
pbmc.markers %>%
  group_by(cluster) %>%
  slice_max(n = 2, order_by = avg_log2FC)
```

Visualize marker expression.

```{r}
FeaturePlot(pbmc, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A", "FCGR3A", "LYZ", "PPBP", "CD8A"))
pbmc.markers %>%
  group_by(cluster) %>%
  top_n(n = 10, wt = avg_log2FC) -> top10
DoHeatmap(pbmc, features = top10$gene) + NoLegend()
```

## Cell Type Assignment

Assign cell types based on known markers.

```{r}
new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B", "CD8 T", "FCGR3A+ Mono",
                     "NK", "DC", "Platelet")
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)
DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```

## Gene-Gene and Cell-Cell Correlation

Calculate correlations for network analysis.

```{r}
M_gg <- cor(t(as.data.frame(pbmc.data[top10$gene,])))
ComplexHeatmap::Heatmap(M_gg, column_names_gp = grid::gpar(fontsize = 6),
                        row_names_gp = grid::gpar(fontsize = 6))
M_cc <- cor(as.data.frame(pbmc.data[top10$gene, 1:400]))
ComplexHeatmap::Heatmap(M_cc, column_names_gp = grid::gpar(fontsize = 6),
                        row_names_gp = grid::gpar(fontsize = 6))
```

## Resources

- [Analysis of Single-Cell RNA-Seq Data](https://broadinstitute.github.io/2020_scWorkshop/index.html)
- [Seurat R Toolkit](https://satijalab.org/seurat/)
- [Scanpy – Single-Cell Analysis in Python](https://scanpy.readthedocs.io/en/stable/)