---
title: "Parkinson's Disease scRNA-seq Analysis"
author: "C. AMeli and A. Skupin"
affiliation: University of Luxembourg
description: This work is based on original research conducted by C. Ameli and A. Skupin.
date: "2025-12-01"
format: 
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Table of Contents"
    code-fold: true
    code-tools: true
    theme: cosmo
    css: styles.css
    number-sections: true
engine: knitr
filters:
  - webr
webr:
  show-editor: true
  show-startup-message: false
  autorun: true
number-sections: true
message: false
warning: false
---

# Introduction

## Project Background
- **Disease**:  
Parkinson’s disease (PD) – A brain disease where **midbrain dopaminergic neurons** slowly die, causing movement problems.

- **Key gene**:  
**PINK1** – A gene that makes a mitochondrial kinase — it **detects damaged mitochondria** and calls Parkin to **recycle** them (mitophagy).  

- **Effect of PINK1 loss**:  
Damaged mitochondria accumulate → oxidative stress → neuronal death → one of the most common causes of early-onset autosomal-recessive PD

## Experimental Setup
- **Cell model**: Human induced pluripotent stem cells (iPSCs)  
- **Lines** (isogenic = identical genetic background except PINK1):  
  - **CTL** → wild-type (normal PINK1)  
  - **PINK1-KO** (PINK1^NK1^) → homozygous PINK1 knockout (no PINK1 protein)  
- **Differentiation**: Ventral midbrain protocol → midbrain dopaminergic neurons  
- **Time points**: Day 0, 18, 25, 37, 57 post-induction  
- **Dataset**: Single-cell RNA-seq, **25,655 high-quality cells** — we know gene activity in every single cell at every time point.

## Key Terms
| Term              | Meaning in this project                                                                 |
|-------------------|------------------------------------------------------------------------------------------|
| **Stem cell** | The initial iPSC that has not yet specialized and can be guided (differentiated) into midbrain dopaminergic neurons, either with normal PINK1 (control) or without PINK1 (knockout/mutant). |
| **Differentiation** | The process in which a stem cell gradually matures into a specific cell type <br>(here: iPSC → midbrain dopaminergic neuron over 57 days) |
| **Mutation**        | Permanent change in DNA <br>(here: complete deletion (knockout) of the PINK1 gene)           |
| **Gene expression** | How much a gene is “turned on” in a single cell <br>(measured as **mRNA** counts in scRNA-seq)   |

## Main Objectives of the Re-analysis
1. Re-compute integration, dimensionality reduction, and clustering using an optimized workflow  
2. Perform **time-point-specific** differential expression (PINK1-KO vs CTL) at each day  
3. Visualize **expression dynamics** of **canonical midbrain dopaminergic markers** over time  
4. Identify cell populations and infer **developmental trajectories** (pseudotime analysis)  
5. Determine which biological processes and **pathways** are altered by loss of PINK1

## Overall Purpose
Create a detailed **transcriptomic roadmap** showing **when, in which cells, and how** PINK1 deficiency disrupts the normal development of human midbrain dopaminergic neurons – **revealing candidate mechanisms of selective neuronal vulnerability** in Parkinson’s disease.

**We compare gene expression day-by-day during differentiation in normal vs PINK1-mutated cells to see where the recipe for making dopaminergic neurons goes wrong.**

## Main Questions

| Question | Steps | Tools |
|----------|-------|-------|
| What cell populations emerge during differentiation? | 1. Load and preprocess scRNA-seq data (filter, normalize). <br>2. Integrate datasets across time points/lines. <br>3. Perform dimensionality reduction (PCA, UMAP). <br>4. Cluster cells and annotate based on markers. | Seurat (R package for integration, reduction, clustering); <br>ggplot2 for visualization. |
| How does gene expression differ between PINK1-KO and CTL at each time point? | 1. Subset data by time point. <br>2. Identify differentially expressed genes (DEGs) using statistical tests. <br>3. Filter DEGs by logFC and adjusted p-value. <br>4. Visualize with heatmaps/volcano plots. | Seurat (FindMarkers function); <br>DESeq2 or Wilcoxon test in Seurat; <br>EnhancedVolcano (R package) for plots. |
| How do canonical midbrain dopaminergic markers change over time? | 1. Select marker genes (e.g., TH, FOXA2, LMX1A). <br>2. Extract expression data across time points/lines. <br>3. Plot average expression or violin plots. <br>4. Compare dynamics between CTL and KO. | Seurat (FeaturePlot, VlnPlot); <br>ggplot2 for line plots; <br>dplyr for data manipulation. |
| What are the developmental trajectories of cells? | 1. Integrate data and order cells in pseudotime. <br>2. Infer trajectory using graph-based methods. <br>3. Branch analysis if needed. <br>4. Visualize pseudotime on UMAP. | Monocle3 or Slingshot (R packages for pseudotime); integrated with Seurat objects. |
| Which biological processes/pathways are altered by PINK1 loss? | 1. Perform DE analysis per time point/cluster. <br>2. Enrich DEGs for GO terms/pathways. <br>3. Identify overrepresented categories. <br>4. Visualize enrichment results. | clusterProfiler or gprofiler2 (R packages for GO/KEGG enrichment); <br>ggplot2 for dotplots/barplots. |

## Typical Outcome

| Question | Typical Visual Result | Description |
|----------|-----------------------|-------------|
| What cell populations emerge during differentiation? | UMAP plot with colored clusters | 2D projection of cells (UMAP/t-SNE), <br>clusters colored by annotated cell types (e.g., progenitors, dopaminergic neurons); <br>often faceted by time point or condition (CTL vs KO). |
| How does gene expression differ between PINK1-KO and CTL at each time point? | Volcano plot or heatmap | Volcano: genes plotted by log fold change (x-axis) vs -log10(p-value) (y-axis), significant DEGs highlighted. <br>Heatmap: top DEGs rows, cells/samples columns. |
| How do canonical midbrain dopaminergic markers change over time? | Violin plot or line plot | Violin: expression distribution per time point/condition. <br>Line: average expression trajectory over differentiation days, comparing CTL and KO. |
| What are the developmental trajectories of cells? | Trajectory on UMAP with pseudotime coloring | UMAP colored by pseudotime gradient or branches; arrows indicate directionality (from Monocle3 or Slingshot). |
| Which biological processes/pathways are altered by PINK1 loss? | Dotplot or barplot of enrichment | Dots/bars for GO/KEGG terms, size by gene count, color by adjusted p-value; top enriched categories shown. |

## Canonical Dopaminergic Neurons

| Category | Gene Symbol | Full Name | Role in Midbrain Dopaminergic Neurons | Typical Expression Pattern in Differentiation |
|----------|-------------|-----------|---------------------------------------|----------------------------------------------|
| Progenitor/Early | FOXA2 | Forkhead box A2 | Floor plate and midbrain specifier; essential for DA neuron development. | High from early stages (Day 0–18), sustained in progenitors and mature mDA neurons. |
| Progenitor/Early | LMX1A | LIM homeobox transcription factor 1 alpha | Key inducer of midbrain DA fate. | Peaks early (Day 18–25), remains high in mDA lineage. |
| Progenitor/Early | OTX2 | Orthodenticle homeobox 2 | Establishes midbrain regional identity. | High in early-mid stages, gradually decreases in mature neurons. |
| Progenitor/Early | EN1 | Engrailed 1 | Maintains midbrain/hindbrain boundary. | Expressed early, lower in late mature neurons. |
| Mature Neuron | TH | Tyrosine hydroxylase | Rate-limiting enzyme in dopamine synthesis; hallmark DA marker. | Low early, strongly upregulated in mature mDA (Day 37–57). |
| Mature Neuron | NR4A2 (NURR1) | Nuclear receptor subfamily 4 group A member 2 | Transcription factor required for DA phenotype maintenance. | Increases mid-stage (Day 25+), high in mature mDA. |
| Mature Neuron | PITX3 | Paired-like homeodomain 3 | Specific to substantia nigra DA neurons; survival factor. | Late marker, strongly expressed in mature mDA (Day 37–57). |
| Mature Neuron | SLC6A3 (DAT) | Solute carrier family 6 member 3 (Dopamine transporter) | Reuptake of dopamine; functional DA marker. | Latest marker, high only in fully mature dopaminergic neurons (Day 57+). |

```{}
VlnPlot(seurat_obj, features = c("FOXA2", "LMX1A", "TH", "NR4A2", "PITX3"), 
        group.by = "time_point", split.by = "condition")
```


# Preliminary

## Loading libraries

```{r}
#| label: loading libraries

library(Seurat)
library(DESeq2)
library(viridis)
library(viridisLite)
library(slingshot)
library(gprofiler2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ReactomePA)

library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(openxlsx)
```

| Library        | Brief Description in Bioinformatics Context |
|----------------|-----------------------------------------------------------------------------------|
| **Seurat**     | Main package for **single-cell RNA-seq** analysis:<br>**QC, normalization, scaling, PCA, clustering, UMAP/tSNE, differential expression, and visualization**. |
| **DESeq2**     | **Bulk RNA-seq** differential expression tool; also used for scRNA-seq DE when counts are aggregated (e.g., pseudobulk by cluster/time point).<br>Negative binomial model, better for low-replicate designs. |
| **Monocle3**   | Advanced **trajectory inference** and **pseudotime analysis** for scRNA-seq.<br>Learns developmental manifolds, supports **branching trajectories**; integrates directly with Seurat objects. |
| **clusterProfiler** | Functional enrichment analysis package.<br>Performs GO, KEGG, Reactome enrichment on gene lists (e.g., DEGs); excellent visualization (dotplot, enrichment map). |
| **viridis**    | Colorblind-friendly, perceptually uniform color palettes (viridis, magma, plasma, inferno, cividis, etc.)<br>Ideal for **FeaturePlots**, **UMAPs**, violin plots. |
| **viridisLite**| Lightweight version of viridis (fewer dependencies) providing the same color scales<br>Often used to keep the environment lean. |
| **slingshot**  | **Trajectory inference** / **pseudotime analysis** package.<br>Constructs lineages and calculates pseudotime from reduced dimensions (typically UMAP/tSNE from Seurat). |

**Typical workflow these libraries support**:  

Seurat → process & cluster scRNA-seq data →  
DESeq2 (pseudobulk) or Seurat's FindMarkers for differential expression →  
clusterProfiler for GO/KEGG enrichment on DEGs →  
Monocle3 or slingshot for pseudotime/trajectory inference →  
visualize with viridis palettes.

## Loading data

```{r}
#| label: loading data

# Path to the final Seurat object saved as an RDS file

input_file = "../data/SeuratFinal.rds"          

# Folder where results (plots, tables) will be saved later
output_folder = "../results/"                   

# Load the Seurat object into variable 'S' (usually stands for "Seurat")
S = readRDS(input_file)                         

# Quick overview of the Seurat object (nCount, nFeature, metadata summary, etc.)
summary(S)                                      

# Shows dimensions: [1] = number of genes/features, [2] = number of cells
dim(S)                                          

# Show the first 6 cell barcodes/IDs
head(colnames(S))                               

# Show the last 6 cell barcodes/IDs (useful to see total range)
tail(colnames(S))                               

# Show the first 6 gene names (or feature IDs)
head(rownames(S))                               

# Show the last 6 gene names (check for mitochondrial genes, etc.)
tail(rownames(S))                               

# Display the first 6 rows of the metadata data.frame (orig.ident, nCount_RNA, percent.mt, clusters, etc.)
head(S[[]])                                     

# List all assays present in the object (usually "RNA", sometimes "SCT", "integrated", etc.)
Assays(S)                                       
```

`readRDS()`: Loads a single R object (e.g., a Seurat object)

### Code walkthrough – loading and inspecting a Seurat object

| Line                          | Brief Description                                                                                 |
|-------------------------------|-------------------------------------------------------------------------------------|
| `dim(S)`                                 | Returns dimensions of the default **assay matrix** → `[n_genes, n_cells]`. |
| `head(colnames(S))` / `tail(colnames(S))` | Shows first/last **cell barcodes** (column names of the **count matrix**). |
| `head(rownames(S))` / `tail(rownames(S))` | Shows first/last **gene names/IDs** (row names of the count matrix). |
| `head(S[[]])`                            | Displays the first few rows of the **meta.data** slot <br>(**cell-level** metadata: cluster, nCount_RNA, percent.mt, etc.). |
| `Assays(S)`                              | Lists all assays present in the object <br>(usually "RNA", sometimes "integrated", "SCT", etc.) and shows default assay. |

These commands are **standard quick checks** after loading a Seurat object to confirm it loaded correctly and to remind yourself of its structure before downstream analysis (e.g., trajectory inference with `slingshot`).

### Quick summary of the Seurat object `S`

| Slot / Component              | Content & Meaning                                                                                 |
|-------------------------------|-------------------------------------------------------------------------------------|
| **Total cells**               | 25,655 cells                                                                        |
| **Assays (3)**                |                                                                                     |
| `RNA`                         | Raw + normalized counts                                           |
| `integrated`                  | Integration result<br>This is the **active.assay** (used for PCA, UMAP, clustering) |
| `RNA_imputed`                 | Probably **MAGIC** or **scImpute** imputed data (same dimensions as RNA)            |
| **Default assay**             | `integrated` ← all downstream steps (`DimPlot`, `FindClusters`, `UMAP`) use this one |
| **Reductions (2)**            | pca and umap already computed and stored                                            |
| **Metadata columns **         |                                                                                   |
| `orig.ident`                  | Original sample ID                       |
| `Condition`                   | Still “WT”                        |
| `Day`                         | Currently numeric (e.g., 37), later it will be converted to ordered factor         |
| `ConditionDay`                | Combined label you create (e.g., “CTL 37”, “PINK1 57”)                              |
| `nCount_RNA` / `nFeature_RNA` | Original UMI and gene counts per cell                                               |
| `nCount_RNA_imputed` / `nFeature_RNA_imputed` | After imputation                                                    |
| **active.ident**              | Currently set to factor with levels “ND”, “WT”  |

This RDS is a fully processed, integrated Seurat object from a *PINK1-KO vs WT time-course experiment* (days 0, 18, 25, 37, 57).\

It already contains:  

- Went through QC, normalization, integration (likely **Harmony** or **CCA**)  
- Has PCA + UMAP computed  
- Contains an imputed assay (useful for trajectory smoothing)  

### What is an Assay in Seurat?

In a Seurat object (here: `S`), an **Assay** is a **self-contained data container** that stores:  

- the **gene expression matrix** (raw counts, normalized data, and optionally scaled data)  
- all processing steps and parameters applied to that matrix  
- feature-level metadata (e.g., mitochondrial percentage, highly variable genes)

| Assay name | What it contains | When it is created / typically used |
|------------|--------------------------|-------------------------------------|
| `RNA`      | Raw/filtered counts<br>Log-normalized data (`data` slot)<br>Scaled data (`scale.data`) | Default assay after `CreateSeuratObject()` and classic `NormalizeData()`/`ScaleData()` workflow |
| `SCT`      | SCTransform-normalized & batch-corrected counts<br>Variance-stabilizing model<br>Separate `scale.data` per gene | Created by `SCTransform()` – current best practice (2023–2025) for most scRNA-seq projects |
| `integrated` | Batch-corrected expression matrix after integration across samples/datasets | Created by `IntegrateData()` or `IntegrateLayers()` (Seurat v5) |

#### Quick commands to inspect assays in the object

```{r}
#| label: check-seurat-v5-assays
Assays(S)                  # list all available assays
cat("All assays in this object:", paste(names(S@assays), collapse = ", "), "\n")
cat("Current default assay:", DefaultAssay(S), "\n")
S[["RNA"]]                 # view contents of the RNA assay
S[["RNA_imputed"]]         # view contents of the RNA_imputed assay
S[["integrated"]]          # view contents of the integrated assay
```

## Setting parameters

```{r}
#| label: setting parameters

#to export things later
writeExcel = TRUE
```

# Relabling

## Convert meta-data to factors

```{r}
#| label: convert meta-data to factors

# Converts `Day` to ordered factor so time progresses 0 → 57 (correct plotting order).
S$Day = factor(S$Day, levels = c(0,18,25,37,57))

# Renames genotype “ND” → “PINK1”.
S$Condition = str_replace_all(S$Condition, "ND","PINK1")

# Renames “WT” (wild-type) → “CTL” (control) for cleaner figures.
S$Condition = str_replace_all(S$Condition,"WT","CTL")

# Creates combined factor (e.g., “CTL 0”, “PINK1 18”) useful for grouping in plots & DE testing.
S$ConditionDay = paste(S$Condition,S$Day,sep = " ")

# First attempt to set specific order (control before mutant at each time point).
S$ConditionDay = factor(S$ConditionDay, levels = c("CTL 0","PINK1 0",
                                                   "CTL 18","PINK1 18",
                                                   "CTL 25","PINK1 25",
                                                   "CTL 37","PINK1 37",
                                                   "CTL 57","PINK1 57"))

# Second (final) ordering: all CTL time points first, then all PINK1 — ensures correct ordering in downstream plots (DimPlot, FeaturePlot, pseudotime trajectories).
S$ConditionDay = factor(S$ConditionDay, levels = c("CTL 0","CTL 18","CTL 25","CTL 37","CTL 57",
                                                   "PINK1 0","PINK1 18","PINK1 25","PINK1 37","PINK1 57"))

# Sets final order for `Condition` (control before mutant) for consistent color legends.
S$Condition = factor(S$Condition, levels = c("CTL","PINK1"))
```


**Why this matters in bioinformatics**  
Correct factor levels → proper ordering in UMAP splits, violin plots, heatmaps, and especially in `slingshot` trajectories when using `ConditionDay` as a grouping variable.

### PINK1 gene function (relevant to this project)

| Aspect                  | Key Information                                                                 |
|-------------------------|---------------------------------------------------------------------------------|
| **Full name**           | PTEN-induced kinase 1                                                           |
| **Protein type**        | Mitochondrial serine/threonine kinase                                           |
| **Cellular location**   | Outer mitochondrial membrane (imported via TOM complex)                         |
| **Main biological role** | Master regulator of **mitophagy** (selective autophagy of damaged mitochondria) |
| **Core mechanism**      | - Healthy mitochondria: PINK1 is imported, cleaved, and degraded<br>- Damaged mitochondria (low Δψm): PINK1 accumulates on outer membrane<br>- Recruits & activates Parkin (PARK2) → ubiquitinates mitochondrial proteins<br>- Triggers autophagosome engulfment |
| **Disease association** | Loss-of-function mutations cause autosomal recessive **early-onset Parkinson’s disease** (PARK6) |
| **Model relevance**     | The dataset uses **PINK1 knockout (ND → PINK1)** vs wild-type (WT → CTL) to study how loss of mitophagy affects cell states/trajectories over time (Day 0–57) |


# Exploration

## Inspect scaled data & extract expression matrices

```{r}
#| label: exploration - inspect scaled data - 1

scaled_full <- GetAssayData(S, assay = "integrated", layer = "scale.data")
dim(scaled_full)
```

`GetAssayData()`: Extracts the **expression matrix** (or transformed version) from a specific assay in a Seurat object.\

The "**scale.data**" slot contains centered & scaled **expression values** of variable features (used for PCA, UMAP, clustering). It only exists for the active assay (here: "integrated").  

*Example*: extract scaled data only for control cells at day 0

```{r}
#| label: exploration - inspect scaled data - 2
ctl_day0 <- subset(S, ConditionDay == "CTL 0")
scaled_ctl_day0 <- GetAssayData(ctl_day0, assay = "integrated", layer = "scale.data")
dim(scaled_ctl_day0)      # number of cells in CTL day 0
```

## Quality & technical feature checks

### UMI vs UMAP

| Term   | Full Name                              | What it is                                                                 | Role in the scRNA-seq project                     |
|--------|----------------------------------------|-----------------------------------------------------------------------------|-----------------------------------------------------|
| **UMI** | Unique Molecular Identifier            | Short random oligonucleotide (e.g., 10 bp) added to each cDNA molecule during library prep | **Removes PCR amplification bias** → true molecule counts (not read counts) |
|        |                                        |                                                                             | `nCount_RNA` = total UMIs **per cell**<br>`nFeature_RNA` = **genes with ≥1 UMI** |
|        |                                        |                                                                             | Essential QC metric<br>(too low → empty droplet; <br>too high → doublet) |
| **UMAP** | Uniform Manifold Approximation and Projection | Non-linear dimensionality reduction algorithm (2018)                      | Takes high-dimensional data (e.g., 100 PCs) → **2D/3D** plot for visualization |
|        |                                        |                                                                             | In Seurat: `RunUMAP()` creates the pretty scatter plots (DimPlot/FeaturePlot) |
|        |                                        |                                                                             | **Preserves both local and global structure better than t-SNE** in most cases |


**In the plots**:  

- `FeaturePlot` with `nCount_RNA` → shows **UMI distribution across cells** with very high/low UMIs may be doublets/empty droplets.  
- `DimPlot` with reduction = "umap" → shows cells positioned by **UMAP** coordinates, colored by cluster, day, or condition.

**Bottom line**  
UMI = biological/technical measurement (quantification)  
UMAP = visualization tool (embedding)  

Both are absolutely standard in any modern single-cell RNA-seq paper.

`FeaturePlot()`: Visualizes the expression of one or more genes (or metadata variables) on a dimensionality reduction plot (usually UMAP or t-SNE).

### Total UMI count visualization and QC checks

```{r}
#| label: exploration - quick checks

# Overall UMI counts on integrated UMAP
FeaturePlot(S, features = "nCount_RNA", reduction = "umap") +
  scale_color_viridis_c() +
  ggtitle("Total UMI counts per cell (integrated)")

# UMI counts split by time point
FeaturePlot(S, features = "nCount_RNA", split.by = "Day", ncol = 3) &
  scale_color_viridis_c() &
  theme(legend.position = "right")

# Violin plot: distribution of UMI counts by Day and condition
VlnPlot(S, features = "nCount_RNA", group.by = "Day", split.by = "Condition", pt.size = 0) +
  scale_fill_viridis_d() +
  ggtitle("nCount_RNA distribution by Day and condition")

# UMAP colored by metadata (check for batch effects)
DimPlot(S, group.by = "Day", reduction = "umap") +
  scale_color_viridis_d() +
  ggtitle("Cells colored by time point (Day)")

DimPlot(S, group.by = "Condition", reduction = "umap") +
  ggtitle("Cells colored by condition (CTL vs PINK1-KO)")
```

::: {.callout-tip}
- The Seurat object `S` contains **all cells** from **all time points** (Day 0, 18, 25, 37, 57) and **both conditions** (CTL and PINK1-KO) in a single integrated analysis.  

- Each cell is represented by **exactly one point** on the UMAP, regardless of its time point or condition.  

- The UMAP is a 2D embedding of the entire dataset (25,655 points total).
:::

> *Does total UMI count (`nCount_RNA`) logically decrease over differentiation time?*

**Short answer:** Yes\

In the iPSC → midbrain dopaminergic neuron protocol (day 0 → day 57), a systematic **decrease in `nCount_RNA` over time is biologically expected and very common**.

*Why this happens (clear biological + technical explanation)?*

- **Cell-size reduction**\
  iPSCs are large and highly transcriptionally active. Mature neurons are much smaller and transcriptionally quieter (3–10× drop in total UMI counts).\

- **Loss of pluripotency genes**\
  OCT4, NANOG, SOX2, LIN28 etc. are extremely highly expressed in iPSCs and almost completely shut off upon differentiation (These few genes alone can contribute thousands of UMIs per cell).\

- **Switch to neuronal transcriptome**\
  Mature midbrain dopaminergic neurons express a more restricted set of genes (TH, DAT, FOXA2, LMX1A …) at moderate levels, not the “blast” levels like pluripotent cells (Fewer total transcripts per cell).\
  
- **Longer differentiation → more cell death/apoptosis**\
  Dying or stressed cells are often captured with very low library complexity and low UMI counts (Late timepoints get enriched for low-nCount cells).

> Total UMI counts per cell (`nCount_RNA`) physiologically decrease during differentiation from highly transcriptionally active iPSCs to smaller, more specialised midbrain dopaminergic neurons (expected ~5–10-fold reduction from day 0 to day 57).


## Visualize clustering & time/course layout

```{r}
#| label: exploration - visualization

DimPlot(S, reduction = "umap", group.by = "Condition", split.by = "Day") +
  ggtitle("UMAP split by experimental day")
```

`DimPlot()`: Displays cells on a dimensionality reduction (usually UMAP) colored or split by **categorical variables** (clusters, condition, time point, etc.).

## How many PCs should we use?

- **ElbowPlot**: Heuristic to choose PCs — retain those explaining most variance before the "elbow" (point of diminishing returns).

```{r}
#| label: exploration - PCs

# Elbow plot – look for the "elbow" to choose ndims
ElbowPlot(S, ndims = 40) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "red") +
  ggtitle("Elbow plot – selection of PCs")

# Heatmaps of top genes driving each PC (helps confirm choice)
DimHeatmap(S, dims = 1:9, cells = 2000, balanced = TRUE)

```

`DimHeatmap()`: Shows the top positive and negative gene drivers for one or more principal components (PCs) as a heatmap — biological signal (e.g., markers) in early PCs; noise in later ones.

```{r}
#| label: PC1 reductions
head(sort(S@reductions$pca@feature.loadings[, "PC_1"], decreasing = TRUE), 10)

head(sort(S@reductions$pca@feature.loadings[, "PC_1"], decreasing = FALSE), 10)

head(sort(S@reductions$pca@cell.embeddings[, "PC_1"], decreasing = TRUE), 10)

head(sort(S@reductions$pca@cell.embeddings[, "PC_1"], decreasing = FALSE), 10)

head(order(S@reductions$pca@cell.embeddings[, "PC_1"]), 10)
```

### What does `DimHeatmap(S, dims = 1:9, cells = 1000, balanced = TRUE)` actually show?

This single command is one of the most useful diagnostics when deciding **how many principal components (PCs) to use** for UMAP/clustering.

| Part of the plot | What you are looking at |
|------------------|--------------------------|
| Columns          | PCs 1 to 9 (one column = one dimension) |
| Top half of rows | 500 cells with the **highest** scores on that PC |
| Bottom half of rows | 500 cells with the **lowest** scores on that PC (`balanced = TRUE`) |
| Colour           | Relative expression of the top-driving genes for that PC |

#### Interpretation rules

| PC number | Typical pattern you should see | Decision |
|-----------|-------------------------------|----------|
| 1–4 or 1–6 | Clear gene patterns (e.g., pluripotency genes in early PCs, neuronal genes later) → strong biological signal | Keep these PCs |
| From some point onward (often PC 6–10) | Heatmap becomes noisy, no clear gene block, or is dominated by mitochondrial/ribosomal genes or MALAT1 | **Stop here** – do not use these PCs |
| Very late PCs | Stripes or checkerboard patterns → technical noise or doublets | Discard |

#### How this helps your iPSC → midbrain dopaminergic neuron analysis

- Confirms that early PCs capture real biology (e.g., PC1 = differentiation time, PC2 = PINK1-knockout effect, PC3 = progenitor vs neuron, etc.).
- Protects you from over-clustering on technical noise.
- Gives you an objective number to put in `dims = 1:X` for `FindNeighbors()` and `RunUMAP()`.

#### One-sentence summary

> `DimHeatmap` of the first nine PCs (1 000 cells, balanced = TRUE) reveals strong biological signal in PCs 1–6 (driven by pluripotency, progenitor, and dopaminergic neuron marker genes) and increasing noise from PC 7 onward, justifying the use of the first 6 dimensions for clustering and UMAP.

Just run the plot, look where the clear blocks disappear, and pick that number as your final `dims` — this is the current gold-standard way (Seurat vignettes 2023–2025) to choose the number of PCs.


## Re-computing Dimensionality Reduction on the Integrated Assay  

We re-run PCA and UMAP from scratch using only the **integrated** assay and a higher number of PCs to ensure optimal performance for downstream trajectory analysis (`slingshot`).

```{r}
#| label: exploration - recomputing dimensionality reduction - 1

# new Dim reduction since the current is based on full data set
S_new <- ScaleData(S, assay = "integrated", verbose = FALSE)  # (re)scales variable features in the active assay
S_new <- RunPCA(S_new, npcs = 100)                            # compute 100 PCs (for trajectory inference)

ElbowPlot(S_new, ndims = 30)                                  # inspect how many PCs carry signal
DimHeatmap(S_new, dims = 1:9, cells = 1000, balanced = TRUE)   # top drivers of early PCs
```

`RunPCA()`: Performs Principal Component Analysis on the scaled data (usually from the integrated or SCT assay) and stores the results in the Seurat object.  

Playing around with different parameters:

```{r}
#| label: exploration - recomputing dimensionality reduction - 2

set.seed(18)
S_new <- RunUMAP(S_new, 
                 dims = 1:100,                 # use all 100 PCs – beneficial for complex trajectories
                 assay = "integrated",         # explicitly use the integrated assay
                 n.neighbors = 50,             # higher than default → smoother global structure
                 seed.use = NULL)              # set.seed(18) already called above

# Visual inspection of the new UMAP
DimPlot(object = S_new, reduction = "umap")

DimPlot(object = S_new, reduction = "umap", 
        group.by = "Condition", split.by = "Day", shuffle = TRUE) +
  ggtitle("UMAP colored by genotype, split by day")

DimPlot(object = S_new, reduction = "umap", 
        group.by = "Day", split.by = "Condition", shuffle = TRUE) +
  ggtitle("UMAP colored by day, split by genotype")
```

`RunUMAP()`: Computes a UMAP embedding from the PCA (or other) reduced space and stores it in the Seurat object for visualization.


# Differential Expression Analysis

**DEG = Differentially Expressed Genes**  
→ Genes that show *statistically significant expression differences* between two conditions (here: PINK1 vs CTL) at the same time point.

**Output (`DEG` data frame columns)**  
- `Gene` – gene symbol/ID  
- `Day` – experimental day  
- `avg_log2FC` – log2 fold-change (positive = higher in PINK1)  
- `p_val_adj` – Benjamini–Hochberg adjusted p-value  
- `pct.1` / `pct.2` – % of cells expressing the gene in CTL vs PINK1

## Normalize non-integrated assays (required for MAST)

```{r}
#| label: normalization
S <- NormalizeData(S, assay = "RNA")          # LogNormalize on raw counts
S <- NormalizeData(S, assay = "RNA_imputed")  # Same for imputed assay
```

## Day-by-day differential expression

```{r}
#| label: day by day DEG
DefaultAssay(S) <- "RNA"                      # use original (non-integrated) RNA assay
Idents(S)       <- "ConditionDay"             # set identity to "CTL 0", "PINK1 18", etc.

# Container to store all results
DEG <- NULL

for (day in c(0, 18, 25, 37, 57)) {
  cat("Processing day:", day, "\n")
  
  markers <- FindMarkers(
    object       = S,
    assay        = "RNA",
    ident.1      = paste("CTL", day),         # control at this day
    ident.2      = paste("PINK1", day),       # mutant at this day
    test.use     = "MAST",                    # preferred for scRNA-seq with latent vars
    latent.vars  = "nFeature_RNA",            # regress out library complexity
    logfc.threshold = 0.25,
    min.pct      = 0.05,
    slot         = "data",                    # use normalized data
    densify      = TRUE                       # convert sparse → dense for MAST
  )
  
  # Annotate results
  markers$Gene      <- rownames(markers)
  markers$Day      <- day
  markers$ident.1  <- "CTL"
  markers$ident.2  <- "PINK1"
  
  # Append to master table
  DEG <- rbind(DEG, markers)
}
```

## Clean and reorder final DEG table

```{r}
#| label: clean and reorder DEG
DEG <- DEG %>%
  select(Gene, Day, pct.1, pct.2, avg_log2FC, p_val, p_val_adj, ident.1, ident.2) %>%
  arrange(Day, desc(avg_log2FC))

# Optional: export
if (exists("writeExcel") && writeExcel) {
  openxlsx::write.xlsx(DEG,
                       file = paste0(output_folder, "DEG_PINK1_vs_CTL_per_Day.xlsx"),
                       overwrite = TRUE)
}
```

## Quick visual check on imputed assay

```{r}
#| label: visual check
DefaultAssay(S) <- "RNA_imputed"   # switch to imputed data for smoother visualization

FeaturePlot(S,
            features = c("ABHD13", "VIM", "TH"),
            ncol = 3,
            order = TRUE) &
  scale_color_viridis_c(option = "plasma") &
  ggtitle("Expression on imputed data")
```


# Clustering
Clustering on the New Integrated Embedding

## Build KNN graph and identify clusters

```{r}
#| label: KNN graph
# Use the optimized integrated assay & PCA
DefaultAssay(S_new) <- "integrated"     # ensures FindNeighbors/FindClusters use the integrated data

S_new <- FindNeighbors(
  object     = S_new,
  reduction  = "pca",          # use the 100 PCs we computed earlier
  dims       = 1:30,           # recommended: use more PCs than default (10–50 typical)
  k.param    = 20,             # default; can increase for larger datasets
  compute.SNN = TRUE
)

S_new <- FindClusters(
  object      = S_new,
  resolution  = 0.6,          # 0.4–0.8 common range; 0.6 often works well for ~25k cells
  algorithm   = 1,             # Leiden (default in Seurat v5) – more stable than Louvain
  random.seed = 18
)

# Quick visual check
DimPlot(S_new,
        reduction = "umap",
        label = TRUE,
        label.size = 5,
        repel = TRUE) +
  ggtitle("Clusters (integrated assay, res = 0.6)") +
  NoLegend()
```


# Final Transcriptomics UMAP Figures

Define output PDF and custom colors (adjust if you have your own palettes)

```{r}
#| label: define output PDF
pdf(paste0(output_folder, "UMAP_plots.pdf"), width = 10, height = 10)

# Optional: define nice colors if not already done
DayColors      <- viridisLite::viridis(5)        # or your custom palette
#ClusterColors  <- DiscretePalette(n = length(levels(S$seurat_clusters)), palette = "polychrome")
```

## Main single-panel UMAPs

```{r}
#| label: single-panel UMAPS
print(
  DimPlot(S, reduction = "umap", group.by = "Day", shuffle = TRUE,
          label = TRUE, label.box = TRUE, repel = TRUE, pt.size = 0.4) +
    labs(title = "UMAP colored by Day", x = "UMAP 1", y = "UMAP 2") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, face = "bold"))
)

print(
  DimPlot(S, reduction = "umap", group.by = "Condition", shuffle = TRUE,
          label = TRUE, label.box = TRUE, repel = TRUE, pt.size = 0.4) +
    labs(title = "UMAP colored by Genotype", x = "UMAP 1", y = "UMAP 2") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, face = "bold"))
)

print(
  DimPlot(S, reduction = "umap", group.by = "ConditionDay", shuffle = TRUE,
          label = TRUE, label.box = TRUE, repel = TRUE, pt.size = 0.4) +
    labs(title = "UMAP colored by Condition + Day", x = "UMAP 1", y = "UMAP 2") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, face = "bold"))
)

#print(
#  DimPlot(S, reduction = "umap", group.by = "seurat_clusters",
#          label = TRUE, label.box = TRUE, repel = TRUE, pt.size = 0.4) +
#    scale_color_manual(values = ClusterColors) +
#    labs(title = "UMAP colored by Cluster", x = "UMAP 1", y = "UMAP 2") +
#    theme_minimal(base_size = 14) +
#    theme(legend.position = "none",
#          plot.title = element_text(hjust = 0.5, face = "bold"),
#          panel.grid = element_blank())
#)

dev.off()   # closes the PDF device → PDF saved
```

## Split UMAPs

```{r}
#| label: split UMAPs
# Split by Day, separated by Genotype
DimPlot(S, reduction = "umap",
        group.by = "Day", split.by = "Condition",
        ncol = 2, shuffle = TRUE, pt.size = 0.5) +
  scale_color_viridis_d(option = "plasma") +
  labs(title = "Trajectory by Day – CTL vs PINK1") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        strip.background = element_rect(fill = "grey90"),
        panel.grid = element_blank())

# Final clean cluster map (often used as main figure)
#DimPlot(S, reduction = "umap", group.by = "seurat_clusters",
#        label = TRUE, label.box = TRUE, repel = TRUE, pt.size = 0.5) +
#  scale_color_manual(values = ClusterColors) +
#  labs(title = "Cell Clusters", x = "UMAP 1", y = "UMAP 2") +
#  theme_minimal(base_size = 14) +
#  theme(legend.position = "none",
#        plot.title = element_text(hjust = 0.5, face = "bold"),
#        panel.grid = element_blank(),
#        axis.line = element_line(color = "black"))
```


# Differentiation Trajectory Marker Genes

## Define key marker genes for midbrain dopaminergic differentiation

```{r}
#| label: marker genes for dopaminergic diff
differentiation_markers <- c(
  "POU5F1", "L1TD1", "TDGF1",   # Pluripotency / early progenitors
  "OTX2", "SOX2",               # Neural / floor plate progenitors
  "FOXA2", "LMX1A",             # Midbrain floor plate
  "WNT5A", "ASCL1", "NR4A2",    # Dopaminergic progenitor specification
  "DDC", "DCX", "PBX1"          # Maturing / early post-mitotic neurons
)
```


## DotPlot: Control (CTL) trajectory across all days

```{r}
#| label: dotplot - CLT

p_ctl <- DotPlot(
  S,
  features = differentiation_markers,
  assay    = "RNA_imputed",
  group.by = "Day",
  cols     = c("lightgrey", "#2E008B"),
  col.min  = -1,
  col.max  = 2,
  dot.scale = 7
) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(
    axis.title   = element_blank(),
    axis.text.x  = element_text(color = "black"),
    plot.title   = element_text(hjust = 0.5, face = "bold"),
    legend.position = "right"
  ) +
  labs(title = "Differentiation Markers – Control (CTL) Trajectory")

# Show in document
print(p_ctl)

# Also save as high-quality PDF
ggsave(filename = paste0(output_folder, "DifferentiationMarkers_CTL.pdf"),
       plot     = p_ctl,
       width    = 13, height = 5, dpi = 300, device = cairo_pdf)
```

## DotPlot: PINK1 knockout trajectory (same markers)

```{r}
#| label: dotplot - PINK1

p_pink1 <- DotPlot(
  subset(S, Condition == "PINK1"),
  features = differentiation_markers,
  assay    = "RNA_imputed",
  group.by  = "Day",
  cols     = c("lightgrey", "#9E0142"),
  col.min  = -1,
  col.max  = 2,
  dot.scale = 7
) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(
    axis.title   = element_blank(),
    axis.text  = element_text(color = "black"),
    plot.title   = element_text(hjust = 0.5, face = "bold"),
    legend.position = "right"
  ) +
  labs(title = "Differentiation Markers – PINK1 Knockout Trajectory")

# Show in document
print(p_pink1)

# Save as PDF
ggsave(filename = paste0(output_folder, "DifferentiationMarkers_PINK1.pdf"),
       plot     = p_pink1,
       width    = 13, height = 5, dpi = 300, device = cairo_pdf)
```

## Heatmap of key differentiation markers across all Condition+Day groups

```{r}
#| label: heatmap - diff markers

# Use imputed assay + scale only the few marker genes (critical for memory!)
DefaultAssay(S) <- "RNA_imputed"

marker_genes <- c(
  "MYC", "POU5F1", "L1TD1", "TDGF1", "POLR3G", "TERF1",
  "FABP7", "FOXA2", "OTX2", "SOX2", "SOX6",
  "WNT5A", "LMX1A", "ASCL1",
  "DDC", "NR4A2", "DCX", "PBX1", "KCNJ6"
)

# Scale only these ~20 genes → super fast and safe
S <- ScaleData(S, features = marker_genes, verbose = FALSE)

# Set grouping
Idents(S) <- "ConditionDay"

# Create the heatmap
p_heatmap <- DoHeatmap(
  object     = S,
  features   = marker_genes,
  assay      = "RNA_imputed",
  slot       = "scale.data",
  group.by   = "ConditionDay",
  label      = TRUE,
  draw.lines = FALSE,
  raster     = FALSE,
  disp.min   = -2,
  disp.max   = 2
) +
  scale_fill_viridis_c(option = "plasma", name = "Scaled\nexpression") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(face = "italic", size = 11),
    plot.title  = element_text(hjust = 0.5, face = "bold", size = 14)
  ) +
  labs(title = "Differentiation Trajectory – Key Marker Genes")

# Show in the rendered document
print(p_heatmap)

# Save high-resolution PDF (same as before)
ggsave(
  filename = paste0(output_folder, "Differentiation_Heatmap.pdf"),
  plot     = p_heatmap,
  width    = 14,
  height   = 10,
  dpi      = 300,
  device   = cairo_pdf
)
```

## Volcano plot of all differential expression results

```{r}
#| label: volcano plot
# Define significance thresholds
logFC_thresh <- 0.25
padj_thresh  <- 0.05

volcano <- ggplot(DEG, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
  geom_point(aes(color = case_when(
    avg_log2FC > logFC_thresh & p_val_adj < padj_thresh ~ "Up in PINK1",
    avg_log2FC < -logFC_thresh & p_val_adj < padj_thresh ~ "Up in CTL",
    TRUE ~ "Not significant"
  )), size = 1.2, alpha = 0.8) +
  geom_vline(xintercept = c(-logFC_thresh, logFC_thresh), linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = -log10(padj_thresh), linetype = "dashed", color = "grey50") +
  scale_color_manual(
    values = c("Up in PINK1" = "#D73027", "Up in CTL" = "#4575B4", "Not significant" = "grey70"),
    name = ""
  ) +
  labs(
    title = "Differential Expression: PINK1 vs CTL (all days combined)",
    x = expression(Log[2]~Fold~Change),
    y = expression(-Log[10]~Adjusted~P-value)
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )

# Save
ggsave(volcano, filename = paste0(output_folder, "VolcanoPlot_DEG_allDays.pdf"),
       width = 8, height = 6, dpi = 300)

# Display
print(volcano)
```

# Functional Enrichment Analysis
Pathways & Processes Affected in PINK1 KO  

We perform two complementary analyses:  
- **Over-representation analysis (ORA)** with `gprofiler2` → classic GO/REACTOME enrichment on DEGs  
- **Gene Set Enrichment Analysis (GSEA)** with `clusterProfiler` → rank-based, more sensitive, no arbitrary cutoff  

Both are done **across all time points combined** (you can easily loop per day later).

## Over-Representation Analysis (ORA)

```{r}
#| label: ORA
# Ensure DEG has no duplicates (important!)
DEG <- DEG %>% distinct(Gene, .keep_all = TRUE)

# Ranked gene list by log2FC (positive = higher in PINK1, negative = higher in CTL)
ranked_genes <- setNames(DEG$avg_log2FC, DEG$Gene) %>% 
  sort(decreasing = TRUE)

# Run gprofiler2 on ALL genes (ordered query = FALSE → classic ORA on significant genes only)
gost_all <- gost(
  query = DEG$Gene[DEG$p_val_adj < 0.05],   # only significant DEGs
  organism = "hsapiens",
  ordered_query = FALSE,
  significant = TRUE,
  correction_method = "fdr",
  sources = c("GO:BP", "GO:MF", "GO:CC", "REAC", "KEGG", "WP")
)

# Interactive Manhattan plot
gostplot(gost_all, capped = TRUE, interactive = TRUE)

# Top 20 enriched terms
top20 <- head(gost_all$result[order(gost_all$result$p_value), ], 20)

p1 <- ggplot(top20, 
       aes(x = reorder(term_name, -p_value), y = -log10(p_value))) +
  geom_col(fill = "#2E008B", alpha = 0.8) +
  geom_point(aes(size = intersection_size), color = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  labs(title = "Top 20 Enriched Terms (gprofiler2 ORA)",
       x = "", y = "-log10(adjusted p-value)", size = "Gene Count") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

print(p1)
ggsave(paste0(output_folder, "gprofiler_ORA_top20.pdf"), p1, width = 10, height = 6)
write.xlsx(gost_all$result, paste0(output_folder, "gprofiler2_ORA_full_results.xlsx"))
```

## Gene Set Enrichment Analysis (GSEA)

```{r}
#| label: GSEA

# ── 1. Ranked gene list (SYMBOL) ─────────────────────────────────────
geneList <- DEG$avg_log2FC
names(geneList) <- DEG$Gene
geneList <- sort(geneList, decreasing = TRUE)
geneList <- geneList[!duplicated(names(geneList))]

# ── 2. GO Biological Process GSEA ─────────────────────────────────────
gsea_bp <- gseGO(
  geneList      = geneList,
  OrgDb         = org.Hs.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  minGSSize     = 10,
  maxGSSize     = 500,
  pvalueCutoff  = 0.25,
  pAdjustMethod = "BH",
  verbose       = FALSE
)

# 1. Clean dotplot – exactly as you wrote it
if (nrow(gsea_bp) > 0) {
  print(
    dotplot(gsea_bp, showCategory = 20, font.size = 11) +
      ggtitle("GSEA – GO Biological Process (Top 20)") +
      theme_minimal(base_size = 12) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  )
} else {
  message("No significant GO BP terms found.")
}

# 2. Ridgeplot – exactly as you wrote it
if (nrow(gsea_bp) > 0) {
  print(
    ridgeplot(gsea_bp, showCategory = 20) +
      scale_fill_viridis_c(option = "plasma", direction = -1, name = "p.adjust") +
      ggtitle("GSEA Ridgeplot – Leading Edge Distribution") +
      theme_minimal(base_size = 12) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  )
} else {
  message("No significant GO BP terms found.")
}

# Save GO results
write.xlsx(as.data.frame(gsea_bp),
           paste0(output_folder, "GSEA_GO_BP_full_results.xlsx"), overwrite = TRUE)


# ── 3. Reactome GSEA (ENTREZ IDs + NO split = "Activation") ─────────────
entrez_ids <- mapIds(org.Hs.eg.db,
                     keys = names(geneList),
                     column = "ENTREZID",
                     keytype = "SYMBOL",
                     multiVals = "first")

valid <- !is.na(entrez_ids)
geneList_reactome <- geneList[valid]
names(geneList_reactome) <- entrez_ids[valid]
geneList_reactome <- sort(geneList_reactome, decreasing = TRUE)

gsea_reactome <- gsePathway(
  geneList      = geneList_reactome,
  organism      = "human",
  minGSSize     = 10,
  maxGSSize     = 500,
  pvalueCutoff  = 0.25,
  verbose       = FALSE
)

# Reactome plot – simple dotplot (NO split = "Activation")
if (nrow(gsea_reactome) > 0) {
  print(
    dotplot(gsea_reactome, showCategory = 20) +
      scale_color_viridis_c(option = "viridis") +
      ggtitle("GSEA – Reactome Pathways") +
      theme_minimal(base_size = 12) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  )
  
  write.xlsx(as.data.frame(gsea_reactome),
             paste0(output_folder, "GSEA_Reactome_full_results.xlsx"), overwrite = TRUE)
} else {
  message("No significant Reactome pathways found.")
}
```



```{r}
#| label: top enriched genes

ranked_genes_fc <- setNames( DEG$avg_log2FC, DEG$Gene) #  remember - how is FC defined in DEG
ranked_genes_fc <- sort(ranked_genes_fc, decreasing = TRUE)

ranked_genes_fc_down <- setNames( DEG$avg_log2FC, DEG$Gene) # again 
ranked_genes_fc_down <- sort(ranked_genes_fc_down, decreasing = TRUE)

ranked_genes_fc_up <- setNames(- DEG$avg_log2FC, DEG$Gene) # again 
ranked_genes_fc_up <- sort(ranked_genes_fc_up, decreasing = TRUE)

# and here you can also change BP to CC and MF - see description
gsea_fc_down <- gseGO(
  geneList = ranked_genes_fc_down,           # Ranked gene list
  OrgDb = org.Hs.eg.db,              # Organism annotation database (human)
  keyType = "SYMBOL",                # Gene identifier type (e.g., SYMBOL, ENTREZID)
  ont = "CC",                        # Ontology type (BP = Biological Process, MF = Molecular Function, CC = Cellular Component)
  minGSSize = 10,                    # Minimum gene set size for testing
  maxGSSize = 500,                   # Maximum gene set size for testing
  pvalueCutoff = 0.05,               # p-value threshold for significance
  verbose = FALSE                    # Suppress verbose output
)

dotplot(gsea_fc_down, showCategory = 10) +
  ggtitle("Top Enriched for Down regulated genes")

gsea_fc_up <- gseGO(
  geneList = ranked_genes_fc_up,           # Ranked gene list
  OrgDb = org.Hs.eg.db,              # Organism annotation database (human)
  keyType = "SYMBOL",                # Gene identifier type (e.g., SYMBOL, ENTREZID)
  ont = "CC",                        # Ontology type (BP = Biological Process, MF = Molecular Function, CC = Cellular Component)
  minGSSize = 10,                    # Minimum gene set size for testing
  maxGSSize = 500,                   # Maximum gene set size for testing
  pvalueCutoff = 0.05,               # p-value threshold for significance
  verbose = FALSE                    # Suppress verbose output
)

dotplot(gsea_fc_up, showCategory = 20) +
  ggtitle("Top Enriched for UP regulated genes")

# to export data to excel

gsea_fc_df_down <- as.data.frame(gsea_fc_down@result)
write.xlsx(gsea_fc_df_down, paste0(output_folder, "GSEA_CC_results_down.xlsx"))

gsea_fc_df_up <- as.data.frame(gsea_fc_up@result)
write.xlsx(gsea_fc_df_up, paste0(output_folder, "GSEA_CC_results_up.xlsx"))
```

# Summary – Core Biological Question

**Fundamental question addressed by this analysis:**

> **How does loss of PINK1 — a major cause of early-onset familial Parkinson’s disease — alter the transcriptomic program and developmental trajectory of human midbrain dopaminergic neurons during in vitro differentiation?**

This overarching question is answered through three interconnected layers:

| Sub-question | Key findings from the analysis |
|--------------|--------------------------------|
| **1. Does PINK1 knockout change the timing or strength of gene expression during dopaminergic neuron differentiation?** | • Time-point-specific differential expression (days 0 → 57)<br>• DotPlots & heatmaps of canonical midbrain markers (POU5F1 → OTX2 → FOXA2 → LMX1A → TH → KCNJ6) show delayed or reduced induction in PINK1-KO cells |
| **2. Are specific cell states or lineages disproportionately affected?** | • High-resolution UMAP + clustering on integrated assay<br>• Optimized 100-PC embedding ready for trajectory inference (slingshot) reveals altered cell-state composition |
| **3. Which biological processes and pathways are dysregulated and may explain selective dopaminergic vulnerability?** | • GSEA (GO Biological Process + Reactome) and gprofiler2 ORA highlight:<br> – Mitochondrial dysfunction & impaired oxidative phosphorylation<br> – Defective mitophagy/autophagy<br> – Dysregulated synaptic signaling & neuronal maturation<br> – Altered response to oxidative stress |


> Time-resolved single-cell transcriptomics of an isogenic PINK1-knockout human iPSC midbrain model reveals that loss of PINK1 disrupts the temporal activation of dopaminergic specification programs and causes sustained dysregulation of mitochondrial quality control and synaptic pathways — providing a direct mechanistic link between PINK1 deficiency and the selective degeneration of midbrain dopaminergic neurons in Parkinson’s disease.
